---
title: "Mexico City Pollution Data"
author: "Diego Valle-Jones"
date: "March 18, 2016"
output: 
 github_document:
          toc: true
          fig_width: 8
          fig_height: 5
---

Master: [![Travis-CI Build Status](https://travis-ci.org/diegovalle/aire.zmvm.svg?branch=master)](https://travis-ci.org/diegovalle/aire.zmvm) 

|              |          |
|--------------|---------------|
| __Author:__ | Diego Valle-Jones |
| __License:__ | [BSD_3](https://opensource.org/licenses/BSD-3-Clause) |
| __Status:__ | alpha |
| __Website:__ | [https://github.com/diegovalle/aire.zmvm](https://github.com/diegovalle/aire.zmvm) |


## What does it do?

This package downloads pollution data for the Mexico City metro area. It can download real-time, daily maximum, minimum, or average for each of the pollution measuring stations or geographical zones in the Zona Metropolitana del Valle de México (greater Mexico City)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

For the moment this package is only available from github. For the development version:

```r
if (!require(devtools)) {
    install.packages("devtools")
}
devtools::install_github('diegovalle/aire.zmvm')
```

## Quick Example

The package consists mainly of three functions: 

* ```get_station_data``` to download data for each of the pollution (and wind and temperature) measuring stations.
* ```get_zone_data``` to download data for each of the 5 geographic zones of Mexico City 
* ```get_latest_data``` to download the latest values for each of the pollution measuring stations.

```{r, fig.show='hold',  message=FALSE}
library("aire.zmvm")
library("dplyr")
library("ggplot2")

o3 <- get_station_data(criterion = "MAXIMOS", # Can be MAXIMOS (daily maximum), 
                                                # MINIMOS (daily minimum), 
                                                # or HORARIOS (hourly average)
                       pollutant = "O3", # "SO2", "CO", "NOX", "NO2", "NO", "O3", 
                                         # "PM10", "PM25", "WSP", "WDR", "TMP", "RH"
                       year = 1986:2016) # The earliest year allowed is 1986
knitr::kable(head(o3))

# Daily max among all base stations
o3_max <- o3 %>% 
  group_by(date) %>% 
  summarise(max = ifelse(all(is.na(value)),
                         NA,
                         base::max(value, na.rm = TRUE))) %>%
  na.omit()

# Plot the daily highest pm10 level with trendline
ggplot(o3_max, 
       aes(date, max, group = 1)) +
  geom_point(color = "black", size = .2, alpha = .4) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 25)) +
  ggtitle("Daily maximum O3 levels") +
  ylab("maximum daily O3 value in ppb") +
  xlab("date") +
  geom_vline(xintercept = as.numeric(as.Date("1989-11-20"))) +
  geom_vline(xintercept = as.numeric(as.Date("2015-07-01"))) +
  annotate("text", label = "hoy no circula", 
           x = as.Date("1987-03-20"),
           y = 480) +
  annotate("text", label = "supreme court ruling", 
           x = as.Date("2011-06-20"),
           y = 350) +
  theme_bw()
```
