---
title: "Mexico City Pollution Data"
author: "Diego Valle-Jones"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
 github_document:
          toc: true
          fig_width: 8
          fig_height: 5
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```


[![Travis-CI Build Status](https://travis-ci.org/diegovalle/aire.zmvm.svg?branch=master)](https://travis-ci.org/diegovalle/aire.zmvm) [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/c7kg6o68exx0lirg?svg=true)](https://ci.appveyor.com/project/diegovalle/aire-zmvm/branch/master) [![Coverage Status](https://img.shields.io/codecov/c/github/diegovalle/aire.zmvm/master.svg)](https://codecov.io/github/diegovalle/aire.zmvm?branch=master)

|                       |                                                       |
|-----------------------|-------------------------------------------------------|
| __Author:__           | Diego Valle-Jones                                     |
| __License:__          | [BSD_3](https://opensource.org/licenses/BSD-3-Clause) |
| __Status:__           | Version 0.5.0 on CRAN                                 |
| __Website:__          | [https://hoyodesmog.diegovalle.net/aire.zmvm.html](https://hoyodesmog.diegovalle.net/aire.zmvm.html) |


## What does it do?

Tools for downloading pollution data for the Mexico City metro area. It can download real-time, daily maximum, minimum, or hourly average data for each of the pollution measuring stations or geographical zones in the Zona Metropolitana del Valle de México (greater Mexico City). It also includes the locations of all the measuring stations and a function to perform inverse distance weighting modified to work with wind direction.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

You can always install the development version from GitHub:

```r
if (!require(devtools)) {
    install.packages("devtools")
}
devtools::install_github('diegovalle/aire.zmvm')
```

To install the most recent package version from CRAN type:

```r
install.packages("aire.zmvm")
library(aire.zmvm)
```
Note that the version on CRAN might not reflect the most recent changes made to this package.

## Quick Example

The package consists mainly of four functions: 

* ```get_station_data``` download data for each of the pollution (and wind and temperature) measuring stations in the original units.
* ```get_station_imeca``` download data for each of the pollution stations in IMECAs
* ```get_zone_imeca``` download data for each of the 5 geographic zones of Mexico City 
* ```get_latest_imeca``` download the latest values in IMECAs for each of the pollution measuring stations.
* ```idw360``` Inverse distance weighting modified to work with degrees

```{r, fig.show='hold',  message=FALSE, fig.width=11}
library("aire.zmvm")
library("dplyr")
library("ggplot2")
library("ggseas")

o3 <- get_station_data(criterion = "MAXIMOS", # Can be one of MAXIMOS (daily maximum), 
                                                # MINIMOS (daily minimum), 
                                                # or HORARIOS (hourly average)
                       pollutant = "O3", # Can be one of "SO2", "CO", "NOX", "NO2", "NO", "O3", 
                                         # "PM10", "PM25", "WSP", "WDR", "TMP", "RH"
                       year = 2009:2017) # A numeric vector, the earliest year allowed is 1986
knitr::kable(head(o3))

# Daily max among all base stations
o3_max <- o3 %>% 
  group_by(date) %>% 
  summarise(max = ifelse(all(is.na(value)),
                         NA,
                         base::max(value, na.rm = TRUE))) %>%
  na.omit()

# ozone values at which a contingencia ambiental was declared
# and the dates during which they were valid
# source: http://www.aire.cdmx.gob.mx/descargas/ultima-hora/calidad-aire/pcaa/pcaa-modificaciones.pdf
contingencia <- data.frame(ppb = c(216, 210, 205, 199, 185, 155, 155),
  start = c(2009, 2009.4973, 2010.4973, 2011.5795, 2012.6052, 2016.291, 2016.4986),
  end = c(2009.4973, 2010.4945, 2011.4945, 2012.6025, 2016.2883, 2016.4959, Inf))
max_daily_df <- tsdf(ts(o3_max$max, start = c(2009,1), frequency = 365.25))
ggplot(max_daily_df,
       aes(x = x, y = y)) + 
  geom_line(colour = "grey75", alpha = .5) +
  stat_rollapplyr(width = 30, align = "right", color = "#01C5D2") +
  #geom_vline(xintercept = 2015 + 183/365) +
  geom_segment(data = contingencia, 
               aes(x=start, y=ppb, xend=end, yend=ppb), color="darkred", linetype = 2)  +
  xlab("date") +
  ylab("parts per billion") +
  ggtitle("Maximum daily ozone concentration and 30 day rolling average", 
          subtitle = "Red lines indicate the values necessary to activate a phase I smog alert. \nData source: SEDEMA")

```



