---
title: "Mexico City Pollution Data"
author: "Diego Valle-Jones"
date: "March 18, 2016"
output: 
 github_document:
          toc: true
          fig_width: 8
          fig_height: 5
---

Master: [![Travis-CI Build Status](https://travis-ci.org/diegovalle/aire.zmvm.svg?branch=master)](https://travis-ci.org/diegovalle/aire.zmvm) 

|              |          |
|--------------|---------------|
| __Author:__ | Diego Valle-Jones |
| __License:__ | [BSD_3](https://opensource.org/licenses/BSD-3-Clause) |
| __Status:__ | alpha |
| __Website:__ | [https://github.com/diegovalle/aire.zmvm](https://github.com/diegovalle/aire.zmvm) |


## What does it do?

This package downloads pollution data for the Mexico City metro area. It can download the daily maximum, minimum, or average for each of the pollution measuring stations or geographical zones in the Zona Metropolitana del Valle de Mexico (greater Mexico City)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

For the moment this package is only available from github. For the development version:

```r
if (!require(devtools)) {
    install.packages("devtools")
}
devtools::install_github('diegovalle/aire.zmvm')
```

## Quick Example

The package consists mainly of two functions: 

* ```get_zone_data``` to download data for each of the 5 geographic zones of Mexico City 
* ```get_station_data``` to download data for each of the pollution (and wind and temperature) measuring stations.

```{r, fig.show='hold',  message=FALSE}
library("aire.zmvm")
library("dplyr")
library("ggplot2")

pm_10 <- get_station_data(criterion = "MAXIMOS", # Can be MAXIMOS (daily maximum), MINIMOS (daily minimum), or HORARIOS (hourly average)
                       pollutant = "PM10", # "SO2", "CO", "NOX", "NO2", "NO", "O3", "PM10", "PM2", "WSP", "WDR", "TMP", "RH"
                       year = 2005:2016) # The earliest year allowed is 2005
knitr::kable(head(pm_10))

# Plot the daily highest pm10 level with trendline
ggplot(pm_10 %>% group_by(date) %>% summarise(max = max(value, na.rm = TRUE)), 
       aes(date, max, group = 1)) +
  geom_line(color = "darkgray", size = .2) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 50), se = FALSE) +
  ggtitle("Daily maximum PM10 levels in ppb") +
  theme_bw()
```
